Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2008-06-21 SCP2008/paper/reply-to-reviews.txt

	We are very grateful to the reviewers for their comments and
so helpful suggestions. We have implemented them all. Below is the
detailed response (omitting suggestions about spelling and grammar,
which we have corrected as recommended).

Reviewer #1:

The summary of our work given by the reviewer is profound; we couldn't
have said better ourselves.

Reviewer #2:

> Two larger changes I suggest are restructuring section 4 and performing
> stronger evaluation of the work.
We have implemented them, as described below.

> Restructuring section 4
> The subsections of section 4 are rather random and disjoint. There
> needs to be a more cohesive plan about what is going on in this
> section. Better yet a concrete example that gets built upon in each
> subsection and motivates the subsection's functionality. Sections 4.1,
> 4.3, 4.5, 4.6 seem to be about defining modules for particular
> aspects. Part of Section 4.2 seems superfluous from the comment
> above. The "DOMAIN" module seems superfluous too: you never need to do
> real operations on values, but only generate code for operations on
> values. Hence "DOMAINL" is what you need and "DOMAIN" is not
> needed. Also I don't think "CONTAINER2D" is ever seen again after
> being introduced in Section 4.2 -- this seems troubling.

We added a paragraph at the end of the container section clarifying
that the `current' container is available to the algorithmic aspects,
under the name of |C|. We describe in more detail the `multi-stage'
nature of instantiating the generic algorithm, pointing out which
module is a part of which.

> Section 4.4 is about programming in OCaml and how to maintain the
> algorithm's state.  Section 4.7 is about putting it all together.

> From my understanding of your work here is an idea to make your paper more
> precise and concise.
>  - Move Section 4.4 into Section 3 because that section seems to deal with
>    low-level programming issues.
Moved.

> - In the intro of Section 4 make a link to the list in Section 2 and say your
>   going to pick some of those aspects and explain how they are implemented -
>   one per subsection.
Done

> - Delete the first half of 4.2.
> - Combine 4.1 with the second half of 4.2, dropping the use of "DOMAIN" and
>   just do everything in terms of "DOMAINL".

It seems DOMAIN is a nice and simple abstraction to build upon later;
it also corresponds to our introductory example in Sec 2. We have
completely re-written Sec 4.2 (which is now Sec 5.2), but decided to 
leave DOMAIN in.

> - Sections 4.3, 4.5, 4.6 then become Sections 4.2, 4.3, 4.4.
> - Section 4.7 could be the last subsection of 4 
> or a new whole section with more detail.

This plan has being followed through. Hopefully the introduction to
(now) Sec 5 describes the cohesive plan of the section.

> Strengthening evaluation
> I understand that program generators are complex programs, that
> explaining them can be difficult and that providing concise and
> meaningful code snippets about them can also be hard. Nevertheless a
> good paper on programming program generators needs to do these
> things. At the end of your paper I thought you had laid down all the
> pieces of the machinery well, but I didn't have much of a sense of how
> the machine runs. So I kind of have to take your word that everything
> works, especially since the code on p24-5 is not explained in detail
> and the appendix code is not readable (by itself it's not very
> convincing: how can I tell this code demonstrates something good?). In
> general, and this would strengthen the merit of the paper, code
> snippets need to be explained better - too often I see a new
> identifier in the code that is not explained.

The comparison of the generated code with the hand-written code
hopefully tells now that the generated code does do the right
thing. We have expanded many explanations for the code snippets,
explicitly relating them to the hand-written sample code -- at the
expense of the length of the paper.

> Showing and explaining the code for the unstaged algorithm might help
> and would give a point of reference to compare the generated code
> with. Explaining the generated code could also be helpful, but maybe
> too hard. Maybe it would be possible to focus on one or more small
> areas of the algorithm and show: this is what the area looks like in
> unstaged program, this is the part of the code generator responsible
> for producing code for this area, and this is the area in the
> generated code. Something like this would give the reader a greater
> understanding of how logic flows through your generation process and
> how your functor technique is ultimately in control of fine-grained
> details of code building.

We have done exactly that. We added Sec 4 with the hand-written
unstaged code, and refer to it when discussing various aspects,
especially the determinant aspect.

> p23, "off-shoring": give relevant citation
Done. Incidentally, Walid Taha spells offshoring without the hyphen,
so we have adjusted the spelling.

> p2, line 1, For an introduction paragraph this could be clearer and stronger.
> What are the "aspects of linear algebra"? How are they addressed and by what?
> (I know they are explained later on.) Why is "may" emphasized?
We re-wrote the paragraph.

> p2, line 6, LU decomposition is a particular case of Gaussian Elimination, not
> vice-versa
We have made that clear all throughout the text (noting that it is
quite common, especially in the naming of library functions, to treat
GE and LU as quasi-synonyms)

> p2, line 10, this paper doesn't not discuss
> hardware-architecture-specific optimization, in fact native code
> generation seems beyond the scope of this paper, I think all you want
> to say is that generative programming allows the programmer to express
> domain-specific optimizations
Yes, we did

> p4, "The design space", this decomposition of the design space is very
> interesting and an important result, but it would be good if this
> section was more compact. E.g., maybe present the aspects in their
> groups straight up instead of two lists.
That is true. However, as this is important, it needs to
be explained in detail, so that all the relevant information is recorded,
in writing. We do consider this identification of aspects as one of
the contributions.

> p10, Figure 1, I don't follow the intuition behind naming the
> operation "retN"; I don't see how the operation is similar to ordinary
> monadic "ret", or what the "N" means; wouldn't a name like "letM" be
> better?
We have added the explanation of the name (on p11 now). Also we chose
retN for consistency with Swadi et al, where RetN was first
introduced.

> p10, Figure 1, showing the type of "retN" and maybe some other values
> would be helpful
Done for ret, retN and bind.

> p10, the type variable 'p and its constraint: could you explain the
> problem with classifiers in more detail? if the type was "type ('s,
> 'w, 'v) monad" what's wrong with having types like "('a state, 'w,
> ('a, 'v) code) monad"?
There is more explanation at the end of p11 and beginning of
p12. There is nothing wrong, in principle, with the type 
"('a state, 'w,  ('a, 'v) code) monad" other than the fact the types
become bigger and bigger; furthermore, only a couple of components of
the type are really needed when giving a type signature.

> p14, line 4, don't you want to assert that "ZpMake" is a "DOMAIN" as was done
> for "IntegerDomain"?
Actually no: we omitted the annotation on purpose, to demonstrate that
we can do it. This is further explained in the text (the end of p18).

> p15, shouldn't the code here appear as the "code type" implementation
> of an "abstract code" module?
> p15, what are "Logic" and "Idx" used for?
> p16, what is the point of the "S" module? why can't "DOMAINL" be
> parameterized by "T" (containing "rep") directly?
> p16, Section 4.2, none of the abstract code combinators were used in
> the second half of this section: was the first half necessary? what is
> the relationship between these two halves that warrants them being in
> the same subsection?  p16, Section 4.2, I understand the purposes of
> abstracting code representation.  Additionally it adds to the
> genericity of the code generator, which is good, but doesn't incur any
> abstraction overhead, which is also good. However for the purposes of
> this paper it is perhaps an abstraction too far: the paper never needs
> to generate anything other than MetaOCaml code and it increases the
> complexity of code snippets. Mentioning that you do abstract code
> representation in the real implementation but elide it for the paper
> might make the paper more succinct and clearer.

The section on abstract code and lifted domain has been completely
re-written, hopefully the above questions are all answered, in
particular about the meaning of Logic and Idx and the merits of
abstracting over interpretations.

> p17, line 4, where does the identifier "rep" come from in this code?
> p17, line -2, what accompanying code?
> p18, line 4, what is the "decl" function?
> p19, showing the types of these functions would be helpful
> p19, line -3, what is the "DETERMINANT" module? it hasn't been introduced yet
> p20, what are "cmonad" and "omonad"?

Sec 5.2 has been completely re-written; hopefully it now answers all
these questions,

> p17, line 4, what is the design decision for making "Dom" a definition
> inside "CONTAINER2D" instead of a parameter of "CONTAINER2D"? For
> readers not familiar with the OCaml module system it might be helpful
> to provide a brief introduction somewhere.
> p17, line -12, now in the module implementation "Dom" is made functor
> parameter, why? and why can you just replace "rep" with "code"?
We added a new paragraph to explain this, the last but one paragraph 
of Sec 5.3.

> p18, although most readers are probably familiar with simple LU
> decomposition, a brief overview that shows how determinant and rank
> are computed in the algorithm with an unstaged code example showing
> the state needed for tracking would be helpful to demonstrate the
> problems stated here
We added a new section, Sec 4.

> p20, shouldn't there be an "answer" method in the state objects?
This is now covered by proper definition of cmonad/omonad.

> p19, line -2, why do you need to track the determinant for integer
> matrices
We explained that in the new Sec 4.

> p20, what is "classif" used for? initially I thought it meant "class
> if", but I just worked out it means "classifier", so this relates to
> an earlier comment, so I think it is important you explain in detail
> why you need to put the classifier in this object.
That is now explained on p11.

> p20, line 5, "e.g., a piece of code...": the type of "lstate" depends on the
> implementation and neither of the implementations shown set "lstate" to "v
> ref".
AbstractDet contains a tdet, which is defined to be v ref
one line above which, by the open statement right above it, it Dom.v ref.

> p21, keep the names of the sign and magnitude(?) state consistent
> across "decl", "upd_sign" and "fin"; also keep the code to retrieve
> the sign and magnitude state consistent across "upd_sign" and "fin"
We have done that.

> p20, what do "acc", "get", "set" do?
Changed.

> p21, what is "wmatrix"? shouldn't this be an instance of "CONTAINER2D"?
It is now defined. We explained its differences from the CONTAINER2D
(the very end of p29).

> p21, "(...,res) monad" what does "..." mean?
We made it more precise.

> p21, the "unitL" in "decl" is superfluous, right? "mo_extend" will
> already return a monadic unit result
Actually no. mo_extend has the result type ('a,unit) monad whereas
unitL has the result type ('a,unit) cmonad, which corresponds to
('a,('c,unit) code) monad. In other words, the former describes an
action that generates no code whereas the latter is an action that
generates code (such as an assignment operation) that has the type
unit. However, we did mean that decl is an action to be executed
solely for its side effects, and it was used as such in the code (its
result was disregarded). So, the type ('a,unit) monad for it is far
better. We have changed the code and the paper to reflect that. There
is no longer `superfluous' unitL. Thank you for making this point.

> p23, what does "GenLA" do?
It is explained now, p30.

> p23, it would be nice to see the module signatures for the other
> features not discussed thus far
We added a new section, Sec 5.5, with the description of the other
features and their signatures.  Perhaps the code can be moved to an
appendix though?

> p24, missing "perform" in "init"
Actually, it is on the first line.

> p24, "module O = F.Output(F)" the functor "Output" takes a "OUTPUTDEP" not a
> "FEATURES", right?
Right, we explained that in more detail (see the paragraph before code
on p32).

### 2008-06-23 SCP2008/paper/metamonads.tex
take care of the comments I got from Pouya

A survey \cite{Carette06} of
Gaussian elimination implementations in the industrial package Maple
\cite{Monagan:2001:M7P}
found 6 clearly identifiable aspects and 35 different implementations of the
algorithm, as well as 45 implementations of directly related
algorithms such as LU decomposition, Cholesky decomposition, and so on.  We
could manually write each of these implementations, optimizing for particular
aspects and using cut-and-paste to ``share'' similar pieces of code.
Or we can write a very generic procedure that accounts for
all the aspects with appropriate abstractions \cite{Gruntz:1994:IG,Axiom}. The
abstraction mechanisms however -- be they procedure, method or a
function call -- have a significant cost, especially for
high-performance numerical computing \cite{Carette06}. Eliminating
this abstraction overhead involves either complex analyses or
domain-specific knowledge (or both!)
\cite{Kennedy01Telescoping,Veldhuizen:2004,scp-CohenDGHKP06},
and so we can not rely on a general
purpose compiler to assuredly perform such optimizations.

@@

In this paper, we report on our continued progress \cite{CaretteKiselyov05}%
\footnote{We describe here a new version of our generator
  dealing with the complete LU decomposition algorithm, as well as
  linear solving. We worked out previously missing aspects of in-place
  updates, representing permutation matrices, dealing with augmented 
  input matrix, and back-propagation. We have changed the representation
  of domain-specific knowledge about permissible compositions of aspects.
  Also included is a careful description of all the aspects involved, as
  well as documenting our development methodology for highly parametric
  scientific software.}
in using code generation for scientific (both numeric and symbolic)
software.  We will use the algorithm family of Gaussian elimination, 
applied to perform LU decomposition 
and linear system solving, as our running examples to demonstrate our
techniques.  Specifically, our contributions are:
\begin{itemize}
    \item Extending a let-insertion, memoizing monad of
      \cite{KiselyovTaha,SwadiMonadic06} for generating control structures
      such as loops and conditionals. The extension is non-trivial
      because of control dependencies and because let-insertion, as we argue,
      is a control effect on its own: for example
      \texttt{let x = exp in \dots} has a different \emph{effect} within a
      conditional branch.
    \item Implementation of the |perform|-notation (patterned after
      the |do|-notation of Haskell) to make monadic code readable.
    \item Use of functors (including higher-order functors) to
      modularize the generator, express aspects (including results of
      various types) and \emph{insure composability of aspects} even
      for aspects that use state and have to be accounted for in many
      places in the generated code.
    \item Encode domain-specific knowledge in the generators so as to
      catch domain-specific instantiation errors at generation
      time.
    \item Provide a thorough classification of the family of Gaussian
    elimination algorithms.
\end{itemize}

@@

Throughout, we assume that the reader is familiar with the basic LU
decomposition algorithm, which factors a invertible matrix $A$ into a unit
lower triangular matrix $L$ and (usually) an upper triangular matrix $U$,
such that $A = LU$.  Pivoting adds a unitary matrix $P$ such
that the factorization is now $A = PLU$.  The case of numeric matrices is well
covered in~\cite{Golub-vanLoan}.  When $A$ is singular, one can still get
a $PLU$ decomposition with $L$ remaining unit lower-triangular. However, 
$U$ is no longer upper triangular but rather ``staggered''
in the upper triangle.

@@

\begin{enumerate}
    \item \textbf{Abstract domain}.  This group includes
      `mathematical' aspects: domain of matrix elements, length
      measure, full division, normalization, zero testing,
      symmetry and other matrix structure.
    \item \textbf{Concrete representation}: choosing data
      structures for domains, containers, permutation
      matrices. This group also includes packing, in-place
      decomposition, and the
      representation of sparse matrices.
    \item \textbf{Interface} of each generated function, including
      the input and output choices, logging and error reporting.
    \item \textbf{Algorithmic Strategy}, such as fraction-free
      updates, pivoting strategies, augmented matrices.
    \item \textbf{Tracking}, of determinant, rank, or pivot, etc.
    \item \textbf{Interpretation}: whether the result is the program or a
      generator that will produce the program.
\end{enumerate}

@@

Let us consider a more complex expression:
\begin{code}
let param_code1 plus one =
  let gen x y = plus (plus y one) (plus x (plus y one)) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
\end{code}
with two occurrences of |plus y one|,
which may be a rather complex computation which we would rather not do
twice. We might be tempted to rely on the compiler's
common-subexpression elimination optimization. When the generated code is
very complex, however, the compiler may overlook common subexpressions.  Or the
subexpressions may occur in an imperative context where the compiler
might not be able to determine whether lifting them is sound. So, being
conservative, the optimizer will leave the duplicates as they are. 
We may attempt to eliminate subexpressions as follows: 
\begin{code}[commandchars=\\\{\}]
let param_code1' plus one =
  let gen x y = let ce = (plus y one) in  plus ce (plus x ce) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
param_code1' plus one
\evalresult{.<fun x_1 -> fun y_2 -> ((y_2 + 1) + (x_1 + (y_2 + 1)))>.}
\end{code}
The result of |param_code1' plus one| still exhibits duplicate
sub-expressions.  This is because our |let|-insertion optimization only saved
the computation at the generating stage.  We need a combinator that
inserts the |let| expression in the generat\emph{ed} code, in a other words a
combinator |letgen| to be used as
\begin{code}
let ce = letgen (plus y one) in plus ce (plus x ce)
\end{code}
yielding code like 
\begin{code}
.<let t = y + 1 in t + (x + t)>.
\end{code}
But that seems impossible because |letgen exp| has to generate
the expression |.<let t = exp in body>.| but |letgen| does not yet
have the |body|. The body needs a temporary identifier |.<t>.|
that is supposed to be the result of |letgen| itself.  Certainly
|letgen| cannot generate only part of a let-expression, without the
|body|, as all generated expressions in MetaOCaml are well-formed and
complete.

@@

\begin{figure}
\begin{code}[numbers=left,numberblanklines=false,fontsize=\scriptsize]
let find_pivot a (n,m) r c =
  let pivot = ref None in               (* ((i,j), pivot_val) option *)
  begin
  for i = r to n-1 do
    for j = c to m-1 do
      let cur = a.(i*m+j) in
      if not (cur == 0) then
        match !pivot with
        | Some (_,oldpivot) ->
            if abs oldpivot > abs cur then
               pivot := Some ((i,j), cur)
        | None -> pivot := Some ((i,j),cur)
  done; done;
  !pivot
  end
let ge = fun a_orig ->
  let r = ref 0 in                    (* current row index, 0-based *)
  let c = ref 0 in                    (* current column index, 0-based *)
  let a = Array.copy (a_orig.arr) in  (* to avoid clobbering A, save it *)
  let m = a_orig.m in                 (* the number of columns *)
  let n = a_orig.n in                 (* the number of rows *)
  let det_sign = ref 1 in             (* Accumulate sign and magnitude *)
  let det_magn = ref 1 in             (*   of the determinant *)
  while !c < m && !r < n do
    (* Look for a pivot *)
    let pivot = find_pivot a (n,m) !r !c in
    let piv_val = (match  pivot with
    | Some ((piv_r, piv_c),piv_val) ->
        if piv_c <> !c then
          begin
            swap_cols a (n,m) !c piv_c;
            det_sign := - !det_sign (* flip the sign of the det *)
          end;
        if piv_r <> !r then
          begin
            swap_rows a (n,m) (!r,!c) piv_r;
            det_sign := - !det_sign (* flip the sign of the det *)
          end;
        Some piv_val
    | None -> None) in
    (* now do the row-reduction over the (r,c)-(n,m) block *)
    (match piv_val with
    | Some a_rc -> begin
        for ii = !r+1 to n-1 do
          let cur = a.(ii*m + !c) in
          if not (cur == 0)  then
            begin
              for j = !c+1 to m-1 do
                (* fraction-free elimination *)
                a.(ii*m+j) <- (a.(ii*m+j) * a_rc - a.(!r*m+j) * cur) / ! det_magn
              done;
              a.(ii*m+ !c) <- 0
            end;
        done;
        det_magn := a_rc;
        r := !r + 1                 (* advance the rank only if pivot > 0*)
      end
    | None -> det_sign := 0);
    c := !c + 1
    done;
    (* Final result *)
    ({arr=a; n=n; m=m},                 (* The matrix now has the U factor*)
    (if      !det_sign = 0 then  0      (* Compute the signed det *)
     else if !det_sign = 1 then  !det_magn 
     else                        (- !det_magn)),
    !r)                                 (* Rank *)
\end{code}
\caption{Fraction-free in-place Gaussian-elimination over an integer
  matrix}\label{fig:GE-algorithm}
\end{figure}
As is typical of textbook presentations, the main algorithm depends on
a separately-defined function |pivot| to find a pivot. Here we use
full-pivoting, i.e. searching the completel as yet unexamined portion of the
matrix, the rectangular block |(r,c)-(n,m)|, for the element with the 
maximum absolute value. The function returns the value of the
pivot thus found, and its location; the return value is an option type with
|None| indicating that all examined elements are zeroes. In the main
algorithm, after we have found the pivot, we swap the current column 
with the pivot column, and swap the current row with the pivot row (if
necessary). 
After the swap, the $a_{rc}$ element of the matrix is the pivot.
Swapping two rows or two columns changes the sign of the determinant.
After the swaps, the algorithm performs so-called row-reduction
over the |(r,c)-(n,m)| block of the matrix.
We implement the \emph{fraction-free} version of the
algorithm\footnote{sometimes also called the Gauss-Bareiss algorithm}
\cite{Bareiss68},
where the division operation in line 50 assuredly divides two integers
with no remainder, which requires the accumulation
of the determinant.

>>>>

A survey \cite{Carette06} of
Gaussian Elimination implementations in the industrial package Maple
\cite{Monagan:2001:M7P}
found 6 clearly identifiable aspects and 35 different implementations of the
algorithm, as well as 45 implementations of directly related
algorithms such as LU decomposition, Cholesky decomposition, and so on.  We
could manually write each of these implementations, optimizing for particular
aspects and using cut-and-paste to ``share'' similar pieces of code.
Or we can write a very generic procedure that accounts for
all the aspects with appropriate abstractions \cite{Gruntz:1994:IG,Axiom}. The
abstraction mechanisms however -- be they procedure, method or a
function call -- have a significant cost, especially for
high-performance numerical computing \cite{Carette06}. Eliminating
this abstraction overhead involves either complex analyses or
domain-specific knowledge (or both!)
\cite{Kennedy01Telescoping,Veldhuizen:2004,scp-CohenDGHKP06},
and so we can not rely on a general
purpose compiler to assuredly perform such optimizations.

@@

In this paper, we report on our continued progress \cite{CaretteKiselyov05}%
\footnote{We describe here a new version of our generator
  dealing with the complete LU decomposition algorithm, as well as
  linear solving. We worked out previously missing aspects of in-place
  updates, representing permutation matrices, dealing with augmented 
  input matrix, and back-propagation. We have changed the representation
  of domain-specific knowledge about permissible compositions of aspects.
  Also included is a careful description of all the aspects involved, as
  well as documenting our development methodology for highly parametric
  scientific software.}
in using code generation for scientific (both numeric and symbolic)
software.  We will use the algorithm family of Gaussian Elimination, 
applied to perform LU decomposition 
and linear system solving, as our running examples to demonstrate our
techniques.  Specifically, our contributions are:
\begin{itemize}
    \item Extending a let-insertion, memoizing monad of
      \cite{KiselyovTaha,SwadiMonadic06} for generating control structures
      such as loops and conditionals. The extension is non-trivial
      because of control dependencies and because let-insertion, as we argue,
      is a control effect on its own: for example
      \texttt{let x = exp in \dots} has a different \emph{effect} within a
      conditional branch.
    \item Implementation of the |perform|-notation (patterned after
      the |do|-notation of Haskell) to make monadic code readable.
    \item Use of functors (including higher-order functors) to
      modularize the generator, express aspects (including results of
      various types) and \emph{insure composability of aspects} even
      for aspects that use state and have to be accounted for in many
      places in the generated code.
    \item Encode domain-specific knowledge in the generators so as to
      catch domain-specific instantiation errors at generation
      time.
    \item Provide a thorough classification of the family of Gaussian
    Elimination algorithms.
\end{itemize}

@@

Throughout, we assume that the reader is familiar with the basic LU
decomposition algorithm, which factors an invertible matrix $A$ into a unit
lower triangular matrix $L$ and (usually) an upper triangular matrix $U$,
such that $A = LU$.  Pivoting adds a unitary matrix $P$ such
that the factorization is now $A = PLU$.  The case of numeric matrices is well
covered in~\cite{Golub-vanLoan}.  When $A$ is singular, one can still get
a $PLU$ decomposition with $L$ remaining unit lower-triangular. However, 
$U$ is no longer upper triangular but rather ``staggered''
in the upper triangle.

@@

\begin{enumerate}
    \item \textbf{Abstract domain}.  This group includes
      `mathematical' aspects: domain of matrix elements, length
      measure, full division, zero testing,
      symmetry and other matrix structure.
    \item \textbf{Concrete representation}: choosing data
      structures for domains, containers, permutation
      matrices. This group also includes packing, in-place
      decomposition, normalization, and the
      representation of sparse matrices.
    \item \textbf{Interface} of each generated function, including
      the input and output choices, logging and error reporting.
    \item \textbf{Algorithmic Strategy}, such as fraction-free
      updates, pivoting strategies, augmented matrices.
    \item \textbf{Tracking}, of determinant, rank, or pivot, etc.
    \item \textbf{Interpretation}: whether the result is the program or a
      generator that will produce the program.
\end{enumerate}

@@

Let us consider a more complex expression:
\begin{code}
let param_code1 plus one =
  let gen x y = plus (plus y one) (plus x (plus y one)) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
\end{code}
with two occurrences of |plus y one|,
which may be a rather complex computation which we would rather not do
twice. We might be tempted to rely on the compiler's
common-subexpression elimination optimization. When the generated code is
very complex, however, the compiler may overlook common subexpressions.  Or the
subexpressions may occur in an imperative context where the compiler
might not be able to determine whether lifting them is sound. So, being
conservative, the optimizer will leave the duplicates as they are. 
We may attempt to eliminate subexpressions as follows: 
\begin{code}[commandchars=\\\{\}]
let param_code1' plus one =
  let gen x y = let ce = (plus y one) in  plus ce (plus x ce) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
param_code1' plus one
\evalresult{.<fun x_1 -> fun y_2 -> ((y_2 + 1) + (x_1 + (y_2 + 1)))>.}
\end{code}
The result of |param_code1' plus one| still exhibits duplicate
sub-expressions.  This is because our |let|-insertion optimization only saved
the computation at the generating stage.  We need a combinator that
inserts the |let| expression in the generat\emph{ed} code, in other words a
combinator |letgen| to be used as
\begin{code}
let ce = letgen (plus y one) in plus ce (plus x ce)
\end{code}
yielding code like 
\begin{code}
.<let t = y + 1 in t + (x + t)>.
\end{code}
But that seems impossible because |letgen exp| has to generate
the expression |.<let t = exp in body>.| but |letgen| does not yet
have the |body|. The body needs a temporary identifier |.<t>.|
that is supposed to be the result of |letgen| itself.  Certainly
|letgen| cannot generate only part of a let-expression, without the
|body|, as all generated expressions in MetaOCaml are well-formed and
complete.

@@

\begin{figure}
\begin{code}[numbers=left,numberblanklines=false,fontsize=\scriptsize]
let find_pivot a (n,m) r c =
  let pivot = ref None in               (* ((i,j), pivot_val) option *)
  begin
  for i = r to n-1 do
    for j = c to m-1 do
      let cur = a.(i*m+j) in
      if not (cur == 0) then
        match !pivot with
        | Some (_,oldpivot) ->
            if abs oldpivot > abs cur then
               pivot := Some ((i,j), cur)
        | None -> pivot := Some ((i,j),cur)
  done; done;
  !pivot
  end
let ge = fun a_orig ->
  let r = ref 0 in                    (* current row index, 0-based *)
  let c = ref 0 in                    (* current column index, 0-based *)
  let a = Array.copy (a_orig.arr) in  (* to avoid clobbering A, save it *)
  let m = a_orig.m in                 (* the number of columns *)
  let n = a_orig.n in                 (* the number of rows *)
  let det_sign = ref 1 in             (* Accumulate sign and magnitude *)
  let det_magn = ref 1 in             (*   of the determinant *)
  while !c < m && !r < n do
    (* Look for a pivot *)
    let pivot = find_pivot a (n,m) !r !c in
    let piv_val = (match  pivot with
    | Some ((piv_r, piv_c),piv_val) ->
        if piv_c <> !c then
          begin
            swap_cols a (n,m) !c piv_c;
            det_sign := - !det_sign (* flip the sign of the det *)
          end;
        if piv_r <> !r then
          begin
            swap_rows a (n,m) (!r,!c) piv_r;
            det_sign := - !det_sign (* flip the sign of the det *)
          end;
        Some piv_val
    | None -> None) in
    (* now do the row-reduction over the (r,c)-(n,m) block *)
    (match piv_val with
    | Some a_rc -> begin
        for ii = !r+1 to n-1 do
          let cur = a.(ii*m + !c) in
          if not (cur == 0)  then
            begin
              for j = !c+1 to m-1 do
                (* fraction-free elimination *)
                a.(ii*m+j) <- (a.(ii*m+j) * a_rc - a.(!r*m+j) * cur) / ! det_magn
              done;
              a.(ii*m+ !c) <- 0
            end;
        done;
        det_magn := a_rc;
        r := !r + 1                 (* advance the rank only if pivot > 0*)
      end
    | None -> det_sign := 0);
    c := !c + 1
    done;
    (* Final result *)
    ({arr=a; n=n; m=m},                 (* The matrix now has the U factor*)
    (if      !det_sign = 0 then  0      (* Compute the signed det *)
     else if !det_sign = 1 then  !det_magn 
     else                        (- !det_magn)),
    !r)                                 (* Rank *)
\end{code}
\caption{Fraction-free in-place Gaussian Elimination over an integer
  matrix}\label{fig:GE-algorithm}
\end{figure}
As is typical of textbook presentations, the main algorithm depends on
a separately-defined function |pivot| to find a pivot. Here we use
full-pivoting, i.e. searching the complete as yet unexamined portion of the
matrix, the rectangular block |(r,c)-(n,m)|, for the element with the 
maximum absolute value. The function returns the value of the
pivot thus found, and its location; the return value is an option type with
|None| indicating that all examined elements are zeroes. In the main
algorithm, after we have found the pivot, we swap the current column 
with the pivot column, and swap the current row with the pivot row (if
necessary). 
After the swap, the $a_{rc}$ element of the matrix is the pivot.
Swapping two rows or two columns changes the sign of the determinant.
After the swaps, the algorithm performs so-called row-reduction
over the |(r,c)-(n,m)| block of the matrix.
We implement the \emph{fraction-free} version of the
algorithm\footnote{sometimes also called the Gauss-Bareiss algorithm}
\cite{Bareiss68},
where the division operation in line 50 assuredly divides two integers
with no remainder, which requires the accumulation
of the determinant.
